
\chapter{Conclusion\label{sec:Future-Research}}

In this thesis, we have presented efficient deterministic algorithms
for a number of polynomial matrix computation problems, including
the computation of order basis, minimal nullspace basis, matrix inverse,
column basis, unimodular completion, determinant, Hermite normal form,
rank, and rank profile. The algorithm for kernel basis computation
also immediately gives us a new way to solve linear systems. An existing
efficient deterministic method for solving linear systems was given
by \citet{GSSV2012}. The algorithm for column basis also immediately
allows us to compute matrix GCD, column reduced forms and Popov normal
forms for matrices of any dimension. 

We first gave algorithms for computing a shifted order basis of an
$m\times n$ matrix of power series over a field $\mathbb{K}$ with
$m\le n$. For a given order $\sigma$ and balanced shift $\vec{s}$
the first algorithm determines an order basis with a cost of $O^{\sim}(n^{\omega}a)$
field operations in $\mathbb{K}$, where $a=m\sigma/n.$ We then provided
a method to refine the cost to $O^{\sim}(n^{\omega-1}m\sigma)$. %Here $O^{\sim}$
%is just $O$ with log factors omitted and 
%$\MM\left(n,d\right)$ denotes
%the cost of multiplying two polynomial matrices with dimension $n$
%and degree $d$. 
\begin{comment}
This extends earlier work of Storjohann which only determines a subset
of an order basis that is within a specified degree bound $\delta$
using $O^{\sim}(n^{\omega}\delta)$ field operations for $\delta\ge\lceil m\sigma/n\rceil$.
\end{comment}
{} While the first algorithm addresses the case when the column degrees
of a complete order basis are unbalanced given a balanced input shift,
it is not efficient in the case when an unbalanced shift results in
the row degrees also becoming unbalanced. %The column degrees of a complete basis may be unbalanced, which is
%a major issue we address in the first algorithm. When the input shift
%is unbalanced, the row degrees of the basis can also be unbalanced
%in addition to the unbalanced column degrees. For this, we present
We have presented a second algorithm which balances the high degree
rows and computes an order basis also using $O^{\sim}(n^{\omega}a)$
field operations in the case that the shift is unbalanced but satisfies
the condition $\sum_{i=1}^{n}(\max(\vec{s})-\vec{s}_{i})\le m\sigma$.%
\begin{comment}
Every problem with any unbalanced shift can be in fact reduced to
a problem with a shift that satisfying this condition if the degrees
of a resulting order basis is known. 
\end{comment}
{} %
\begin{comment}
Many unbalanced shift problems can be in fact converted to problems
satisfying this condition. 
\end{comment}
{} This condition essentially allows us to locate those high degree
rows that need to be balanced. %
\begin{comment}
In more general unbalanced shift cases, this algorithm may not work
well directly since we do not know in advance which are the high degree
rows need to be balanced. But it may work efficiently if we have an
effective way of estimating the resulting row degrees. 
\end{comment}
{} %
\begin{comment}
This extends the earlier work by the authors from ISSAC'09.
\end{comment}


We then presented an algorithm for the computation of a minimal nullspace
basis of an $m\times n$ input matrix of univariate polynomials over
a field $\mathbb{K}$ with $m\le n$. This algorithm computes a minimal
nullspace basis of a degree $d$ input matrix with a cost of $O^{\sim}\left(n^{\omega-1}md\right)$
field operations in $\mathbb{K}$. %
\begin{comment}
Here the soft-$O$ notation is Big-$O$ with log factors removed while
$\omega$ is the exponent of matrix multiplication.
\end{comment}
{} The same algorithm also works in the more general situation on computing
a shifted minimal nullspace basis, with a given degree shift $\vec{s}\in\mathbb{Z}^{n}$
whose entries bound the corresponding column degrees of the input
matrix. In this case a $\vec{s}$-minimal right nullspace basis can
be computed with a cost of $O^{\sim}(n^{\omega}s)$ field operations,
where $s$ is the average of the largest $m$ entries of $\vec{s}$. 

Order basis computation and nullspace basis computation were then
applied to the remaining problems. An algorithm for computing the
inverse of an matrix in $\mathbb{K}\left[x\right]^{n\times n}$ was
then given with a cost of $O^{\sim}\left(n^{3}s\right)$ field operations,
where $s$ is the average of the column or row degrees of the input
matrix. The inverse represented alternatively by a product of $\left\lceil \log n\right\rceil $
matrices costs only $O^{\sim}\left(n^{\omega}s\right)$ to compute.
We then discussed the computation of a column basis of an input matrix
in $\mathbb{K}\left[x\right]^{m\times n}$ with a cost of $O^{\sim}\left(m^{\omega}ns\right)$,
where $s$ is again the average column degree of the input matrix.
Next, an algorithm was presented for computing an unimodular completion
of an input matrix in $\mathbb{K}\left[x\right]^{m\times n}$, $m<n$
with a cost of $O^{\sim}\left(n^{\omega}s\right)$, where $s$ is
the average of the $m$ largest column degrees of the input matrix.
Then an algorithm for computing the determinant of an input matrix
in $\mathbb{K}\left[x\right]^{n\times n}$ with a cost of $O^{\sim}\left(n^{\omega}s\right)$
was given, where $s$ is the average column or row degree of the input
matrix. Then we looked at an algorithm for computing the Hermite normal
form of a degree $d$ input matrix in $\mathbb{K}\left[x\right]^{n\times n}$
with a cost of $O^{\sim}\left(n^{\omega}d\right)$. Finally, we provided
algorithms for rank-sensitive computations of the rank and rank profile
of an input matrix in $\mathbb{K}\left[x\right]^{m\times n}$ with
a cost of $O^{\sim}\left(mr^{\omega-2}ns\right)$, where $s$ is the
average column degree of the input matrix, and then applied the rank
profile algorithm to rank-sensitive computation of minimal kernel
basis to obtain a cost of $O^{\sim}(nmr^{\omega-2}d+n^{\omega-1}rd)$.

We reduce all these problems to polynomial matrix multiplications.
The computational costs of our algorithms are then similar to the
costs of multiplying matrices, whose dimensions match the input matrix
dimensions in the original problems, and whose degrees equal the average
column degrees of the original input matrices in most cases. The use
of the average column degrees instead of the commonly used matrix
degrees, or equivalently the maximum column degrees, makes our computational
costs more precise and tighter. In addition, the shifted minimal bases
computed by our algorithms are more general than the standard minimal
base.
