
\chapter{Kernel Basis\label{chap:NullspaceBasis}}

In this chapter we discuss the computation of minimal kernel bases.

Minimal kernel bases can be directly computed via order basis computation.
Indeed if the order $\sigma$ of a $\left(\mathbf{F},\sigma,\vec{s}\right)$-basis
$\mathbf{P}$ is high enough, then $\mathbf{P}$ contains a $\vec{s}$-minimal
kernel basis $\mathbf{N}$. However, this approach may require the
order $\sigma$ to be quite high. For example, if $\mathbf{F}$ has
degree $d$ and $\vec{s}$ is uniform, then its minimal kernel bases
can have degree up to $md$. In that case the order $\sigma$ would
need to be set to $d+md$ in the order basis computation in order
to fully compute a minimal kernel basis. The fastest method of computing
such a $\left(\mathbf{F},d+md\right)$-basis would cost $O^{\sim}\left(n^{\omega-1}m^{2}d\right)$
using the algorithm from \citep{za2009}. We can see from this cost
that there is room for improvement when $m$ is large. For example,
in the worst case when $m\in\Theta\left(n\right)$, this cost would
be $O^{\sim}\left(n^{\omega+1}d\right)$. This points to a root cause
for the inefficiency in this approach. Namely, when $m$ is large,
the computed kernel basis, which can have a column dimension of $n-m$,
is a much smaller subset of the order basis computed. Hence considerable
effort is put in the computation of order basis elements that are
not part of a kernel basis. A key to reducing the cost is therefore
to reduce such computation of unneeded order basis elements, which
is achieved in our algorithm by only using order basis computation
to compute partial kernel bases of low degrees.

\input{nullspaceBasisComputing.tex}
